NumPy---Array Computation for Python

### 2020-01-06 body.tex
Move hybrid development
Move figure

Table X1 about here: backgrounds of initial designers (mainly math, physics).

Table X2 about here: backgrounds of major committers (mainly math, physics).

@@

@@

@@

With its first stable release in October 2006, NumPy unified
the scientific Python community and now underpins almost every library
that does numerical computation, including SciPy\cite{virtanen2019scipy},
pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
and scikit-image\cite{vanderwalt2014scikit}.
NumPy consists of the \code{ndarray}-object along with utility functions that
operate on it.
Because of its inherent simplicity---being a pointer to memory with some
associated meta-data about shape, data-type, and so forth---the NumPy array is
the {\it de facto} exchange format for array data in Python.
The library has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous as
a language for tensor computation---witnessed by its use in popular
deep learning libraries such as PyTorch\cite{pytorch}.

The \code{ndarray} data structure stores regularly spaced elements of a single
data type in a contiguous block of memory; along with meta-data such as shape,
it allows for the efficient representation of $n$-dimensional arrays \cite{vanderwalt2011numpy}.
The need for such a data structure arises often in science.
For instance, they may store measurements from an experiment or simulation
at discrete time points.

Data analysis frequently involves operations $(+, -, \times, /, \log, \sin, \dots)$ on
the individual measurements.
In programming languages such as C or Java, this would involve looping over
every element to perform these operations.
Array computations generalize these element-wise operations on scalars to
vectors (1-d arrays), matrices (2-d arrays), and higher dimensional arrays (n-d
arrays).
For instance, if \code{x} is an \code{ndarray}, then \code{2 * x} produces
an \code{ndarray} whose values are twice those in \code{x}.
Similarly, if \code{x} and \code{y} have the same shape, then
\code{x * y} produces an \code{ndarray} whose values are the products of
the corresponding entries of \code{x} and \code{y}.
These \emph{vectorized} operations often produce one-liners that would be tens
or hundreds of lines of code in languages such as C.

NumPy also provides a large collection of functions for array manipulation
including functions for: creating, reshaping, concatenating, and padding arrays;
searching, sorting and counting data
in arrays; computing elementary statistics, such as the mean, median,
variance, and standard deviation; file I/O; and more.
It also provides extensive support for generating pseudorandom numbers
as well as an assortment of probability distributions.
For historical reasons, NumPy also includes basic functionality for
linear algebra, fast Fourier transforms and windowing,
and polynomial fitting.
We will continue supporting these, but we will not expand beyond them.

With the recent infusion of funding and a clear process for coordinating with
the developer community, we have been able to tackle a number of important
large scale changes.
We highlight two of those below.

@@

To address these issues NumPy has the goal of providing the fundamental
API for \emph{interoperability} between the various NumPy-like APIs.
An earlier step in this direction was the implementation of the
\code{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
for most mathematical functions.\cite{NEP13}
In 2019 this was expanded more generally with the inclusion of the
\code{\_\_array\_function\_\_} protocol into NumPy~1.17.
These two protocols allows providers of array-like APIs to be interoperable
with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
For the users relying on specialized array projects it means that even though
much code is written specifically for NumPy arrays and uses the NumPy API as
\code{import numpy as np}, it can nevertheless work for them.  For
example, here is how a CuPy GPU array can be passed through NumPy for
processing, with all operations being dispatched back to CuPy:

\begin{lstlisting}
import numpy as np
import cupy as cp

x_gpu = cp.array([1, 2, 3])
y = np.sum(x_gpu)  # Returns a GPU array
\end{lstlisting}

@@

The three components are combined to construct a complete random number
generator.

\begin{lstlisting}
from numpy.random import (
    Generator,
    PCG64,
    SeedSequence,
)

seq = SeedSequence(1030424547444117993331016959)
pcg = PCG64(seq)
gen = Generator(pcg)
\end{lstlisting}

\noindent This approach retains access to the seed sequence which can then be
used to spawn additional generators.

\begin{lstlisting}
children = seq.spawn(2)
gen_0 = Generator(PCG64(children[0]))
gen_1 = Generator(PCG64(children[1]))
\end{lstlisting}

\noindent While this approach retains complete flexibility, the method
\code{np.random.default\_rng} can be used to instantiate a \code{Generator} when
reproducibility is not needed.

>>>>

%Table X1 about here: backgrounds of initial designers (mainly math, physics).

%Table X2 about here: backgrounds of major committers (mainly math, physics).

@@

NumPy now underpins almost every Python library that does scientific or
numerical computation, including SciPy\cite{virtanen2019scipy},
matplotlib\cite{hunter2007matplotlib}, pandas\cite{mckinney-proc-scipy-2010},
scikit-learn\cite{pedregosa2011scikit}, and
scikit-image\cite{vanderwalt2014scikit}.
NumPy consists of the $n$-dimensional array object along with utility functions
that operate on it.
Because of its inherent simplicity---being a pointer to memory with some
associated meta-data about shape, data-type, and so forth---the NumPy array is
the {\it de facto} exchange format for array data in Python.
The library has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous as
a language for tensor computation---witnessed by its use in popular
deep learning libraries such as PyTorch\cite{pytorch}.

@@

The NumPy array data structure stores regularly spaced elements of a single
data type in a contiguous block of memory; along with meta-data such as shape,
it allows for the efficient representation of $n$-dimensional arrays \cite{vanderwalt2011numpy}.
The need for such a data structure arises often in science.
For instance, they may store measurements from an experiment or simulation
at discrete time points.

Data analysis frequently involves operations $(+, -, \times, /, \log, \sin, \dots)$ on
the individual measurements.
In programming languages such as C or Java, this would involve looping over
every element to perform these operations.
Array computations generalize these element-wise operations on scalars to
vectors (1-d arrays), matrices (2-d arrays), and higher dimensional arrays (n-d
arrays).
For instance, if \code{x} is an array, then \code{2 * x} produces
an array whose values are twice those in \code{x}.
Similarly, if \code{x} and \code{y} have the same shape, then
\code{x * y} produces an array whose values are the products of
the corresponding entries of \code{x} and \code{y}.
When \code{x} and \code{y} have different shapes, the smaller array is \emph{broadcast}
across the larger array to a common shape when sensible.
These vectorized operations often produce one-liners that would be many tens
of lines of code in languages such as C.

NumPy also provides a large collection of functions for array manipulation
including functions for:
creating, reshaping, concatenating, and padding arrays;
searching, sorting and counting data in arrays;
computing elementary statistics, such as the mean, median, variance, and standard deviation;
reading and writing files;
and more.
It also provides extensive support for generating pseudorandom numbers
as well as an assortment of probability distributions.
For historical reasons, NumPy also includes basic functionality for
linear algebra, fast Fourier transforms and windowing,
and polynomial fitting.
We will continue supporting these, but we will not expand beyond them.

@@

We briefly highlight two of those below.

@@

To address these issues NumPy has the goal of providing the fundamental
API for \emph{interoperability} between the various NumPy-like APIs.
An earlier step in this direction was the implementation of the
\code{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
for most mathematical functions.\cite{NEP13}
In 2019 this was expanded more generally with the inclusion of the
\code{\_\_array\_function\_\_} protocol into NumPy~1.17.
These two protocols allows providers of array-like APIs to be interoperable
with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
For the users relying on specialized array projects it means that even though
much code is written specifically for NumPy arrays and uses the NumPy API as
\code{import numpy as np}, it can nevertheless work for them.
%For example, here is how a CuPy GPU array can be passed through NumPy for
%processing, with all operations being dispatched back to CuPy:
%
%\begin{lstlisting}
%import numpy as np
%import cupy as cp
%
%x_gpu = cp.array([1, 2, 3])
%y = np.sum(x_gpu)  # Returns a GPU array
%\end{lstlisting}

@@

The three components are combined to construct a complete random number
generator.
%
%\begin{lstlisting}
%from numpy.random import (
%    Generator,
%    PCG64,
%    SeedSequence,
%)
%
%seq = SeedSequence(1030424547444117993331016959)
%pcg = PCG64(seq)
%gen = Generator(pcg)
%\end{lstlisting}
%
This approach retains access to the seed sequence which can then be
used to spawn additional generators.
%
%\begin{lstlisting}
%children = seq.spawn(2)
%gen_0 = Generator(PCG64(children[0]))
%gen_1 = Generator(PCG64(children[1]))
%\end{lstlisting}
%
While this approach retains complete flexibility, the method
\code{np.random.default\_rng} can be used to instantiate a \code{Generator} when
reproducibility is not needed.
